Staged Abstract Interpreters

### 2019-02-24 paper/10_intro.tex
misc
typo, monadic ai

\begin{lstlisting}
def power5(b: Int) = {
  val x1 = x0 * x0; val x2 = x0 * x1; val x3 = x0 * x2; val x4 = x0 * x3
  x4
}
\end{lstlisting}

@@

where $e$ is the program of type @Expr@ and @arg@ is the input to program $e$,
and @eval@ produces a value.
Semantically, the interpreter satisfies $ \texttt{eval}(e)(arg) = [\![ e ]\!] arg$ for all
programs and inputs. Given a program $e_0$, by applying the first Futamura
projection, we may obtain a specialized interpreter
$\texttt{eval}_{\texttt{e0}}$ with respect to $e_0$, and by definition of the interpreter, 
$\texttt{eval}_{\texttt{e0}}(arg) = [\![ e_0 ]\!] arg $. If we generalize the
argument @arg@ to an environment (as well as a companion heap object, if for the
interest of using assginments) that contains values mapped from free variables
in $e$, the interpreter is a standard environment/store-passing interpreter,
where the specialization through MSP is also applicable.

Roughly at the same time when Futamura projections was visioned in 1970s,
\citeauthor{DBLP:conf/popl/CousotC77} proposed abstract interpretation as a
lattice-based semantic approach to construct sound static analysis, by
approximation of fixpoints \cite{DBLP:conf/popl/CousotC77}. However,
constructing sound abstract interpreters was considered abstruse and complicated
for a long time.
Recent progress such as Abstracting Abstract Machines (AAM)
uncovers a systematic method to derive sound abstract interpreters from their
concrete counterparts, where the soundness can be easily established by
examining the transformation of semantics.
The AAM approach also has been been applied to different variants of
definitional interpreters and abstract machines \cite{DBLP:journals/jfp/HornM12,
DBLP:conf/icfp/HornM10, DBLP:journals/pacmpl/DaraisLNH17}.
% \todo{CEK example?}

Now we have seen two orthogonal transformations of interpreters: by
speclization, we can refactor interpreters to code generators; by abstraction,
we can refactor interpreters to static analyzers. As an intellectual quest, it
is natural to ask --- starting from an interpreter, \textbf{can we derive a
sound abstract interpreter but it works like a code generator which produces fast
analysis code?} Conceptually, it is similar to the concrete setting: for an
abstract interpreter $\widehat{\texttt{eval}}: \texttt{Expr} \to
\widehat{\texttt{Env}} \to \widehat{\texttt{Value}}$, which takes a program, an abstract
environment and returns abstract values, we would like to specialize it with
repsect to a program $e_0$ and produce an code generator
$\widehat{\texttt{eval}}_{\texttt{e0}} : \widehat{\texttt{Env}} \to
\widehat{\texttt{Value}}$, such that $
\widehat{\texttt{eval}}_{\texttt{e0}}(\widehat{\rho}) =
\widehat{\texttt{eval}}(e_0)(\widehat{\rho})$, but the staged one leaves the
abstract environment $\widehat{\rho}$ dynamic.

@@

In this paper, we study the confluence of two old ideas --- Futamura projection
and abstract interpretation, but from the perspective of their recent
realizations --- multi-stage programming and definitional abstract interpreters.
To be specific, we present the application of the first Futamura projection on
definitional abstract interpreters, which intersects as staged abstract
interpreters. By borrowing the ideas from monadic interpreters
\cite{Steele:1994:BIC:174675.178068, DBLP:conf/popl/LiangHJ95,
DBLP:journals/pacmpl/DaraisLNH17, Sergey:2013:MAI:2491956.2491979} and embedded
domain-specific languages \cite{DBLP:conf/snapl/RompfBLSJAOSKDK15,
DBLP:journals/jfp/CaretteKS09, DBLP:conf/icfp/GibbonsW14}, we develop an
approach to fulfill four different semantics (Figure~\ref{confluence}) in a
unified framework. The four semantics share one same generic interpreter.
However, by instantiating the interpreter differently, it can behave as a
concrete interpreter, an abstracter interperter, or their code-generator
versions, respectively. The key idea that enables this flexibility is to use
abstract type members abstract over concrete/abstract components (e.g., concrete
values or abstract values), as well as the binding-time of them (e.g., static or
dyanmic). By staging, the overhead caused by the monadic layers is eliminated in
the generated code.

>>>>

\begin{lstlisting}
def power5(b: Int) = {
  val x1 = b * b; val x2 = b * x1; val x3 = b * x2; val x4 = b * x3
  x4
}
\end{lstlisting}

@@

where $e$ is the program of type @Expr@ and @arg@ is the input to program $e$,
and @eval@ produces a value.
Semantically, the interpreter satisfies $ \texttt{eval}(e)(arg) = [\![ e ]\!] arg$ for all
programs and inputs. Given a program $e_0$, by applying the first Futamura
projection, we may obtain a specialized interpreter
$\texttt{eval}_{\texttt{e0}}$ with respect to $e_0$, and by definition of the interpreter, 
$\texttt{eval}_{\texttt{e0}}(arg) = [\![ e_0 ]\!] arg $. If we generalize the
argument @arg@ to an environment (as well as a companion heap object, if for the
interest of using assignments) that contains values mapped from free variables
in $e$, the interpreter is a standard environment/store-passing interpreter,
where the specialization through MSP is also applicable.

Roughly at the same time when Futamura projections was visioned in 1970s,
\citeauthor{DBLP:conf/popl/CousotC77} proposed abstract interpretation as a
lattice-based semantic approach to construct sound static analysis, by
approximation of fix-points \cite{DBLP:conf/popl/CousotC77}. However,
constructing sound abstract interpreters was considered abstruse and complicated
for a long time.
Recent progress such as Abstracting Abstract Machines (AAM)
uncovers a systematic method to derive sound abstract interpreters from their
concrete counterparts, where the soundness can be easily established by
examining the transformation of semantics.
The AAM approach also has been been applied to different variants of
definitional interpreters and abstract machines \cite{DBLP:journals/jfp/HornM12,
DBLP:conf/icfp/HornM10, DBLP:journals/pacmpl/DaraisLNH17}.
% \todo{CEK example?}

Now we have seen two orthogonal transformations of interpreters: by
specialization, we can refactor interpreters to code generators; by approximation,
we can refactor interpreters to static analyzers. As an intellectual quest, it
is natural to ask --- starting from an interpreter, \textbf{can we derive a
sound abstract interpreter but it works like a code generator which produces fast
analysis code?} Conceptually, it is similar to the concrete setting: for an
abstract interpreter $\widehat{\texttt{eval}}: \texttt{Expr} \to
\widehat{\texttt{Env}} \to \widehat{\texttt{Value}}$, which takes a program, an abstract
environment and returns abstract values, we would like to specialize it with
respect to a program $e_0$ and produce a compiled analysis
$\widehat{\texttt{eval}}_{\texttt{e0}} : \widehat{\texttt{Env}} \to
\widehat{\texttt{Value}}$, such that $
\widehat{\texttt{eval}}_{\texttt{e0}}(\widehat{\rho}) =
\widehat{\texttt{eval}}(e_0)(\widehat{\rho})$, but the compiled one leaves the
abstract environment $\widehat{\rho}$ as remaining input.

@@

In this paper, we study the confluence of two old ideas --- Futamura projection
and abstract interpretation, but from the perspective of their recent
realizations --- multi-stage programming and definitional abstract interpreters.
To be specific, we present the application of the first Futamura projection on
definitional abstract interpreters, which intersects as staged abstract
interpreters. By borrowing the ideas from monadic interpreters
\cite{Steele:1994:BIC:174675.178068, DBLP:conf/popl/LiangHJ95,
DBLP:journals/pacmpl/DaraisLNH17, Sergey:2013:MAI:2491956.2491979} and embedded
domain-specific languages \cite{DBLP:conf/snapl/RompfBLSJAOSKDK15,
DBLP:journals/jfp/CaretteKS09, DBLP:conf/icfp/GibbonsW14}, we develop an
approach to fulfill four different semantics (Figure~\ref{confluence}) in a
unified framework. The four semantics share one same generic interpreter.
However, by instantiating the interpreter differently, it can behave as a
concrete interpreter, an abstracter interpreter, or their code-generator
versions, respectively. The key idea that enables this flexibility is to use
abstract type members abstract over concrete/abstract components (e.g., concrete
values or abstract values), as well as the binding-time of them (e.g., static or
dynamic). By staging, the overhead caused by the monadic layers is eliminated in
the generated code.
